Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-07-29 write-up/covid-writeup.tex
Updated DR model and latex

\section{Introduction}
The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  \textcolor{red}{As of June 15th, 2021, over 176 million cases have been confirmed worldwide with 33.4 million cases and over 590 thousand confirmed deaths across the United States.} This pandemic
has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.  Public databases~\citep{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.

@@

\subsection{Inverse-probability weighting approach}

@@

Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates. Based on published age-specific IFRs~\cite{Levin2020}, IFR by age seems to closely follow a log-linear relationship. Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
\begin{table}[!th]
\begin{tabular}{c | c c c c c c c c}
Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
\end{tabular}
\caption{IFR by age-strata}
\label{tab:ifrperage}
\end{table}

The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.

Figure~\ref{fig:deathsppc} performs a posterior predictive check on aggregate

\begin{figure}[!th]
\centering
\includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
\caption{Posterior predictive check on aggregate daily death counts}
\label{fig:deathsppc}
\end{figure}

@@

Present IPW1, IPW2, Model-based, DR1, and DR2.

We notice that IPW1/2 don't improve greatly but this may be due to the

We notice model-based rates are much lower, but

\begin{figure}[!th]
\centering
\includegraphics[width=.6\linewidth]{../figs/tv_air.png}
\caption{Active Infection Rate Estimates}
\label{fig:testinglik_mainpaper}
\end{figure}

@@

{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}

There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty. Selection bias leads data analysts to feel certain about incorrect conclusions.  We end with a brief discussion of important related topics.

Length bias and measurement-error?

@@

Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We would argue that, after the current wave, understanding prevalence is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
Random sampling is a powerful tool.  One difficulty with SRS is when prevalence is low.  Consider the Santa Clara study~\citep{Bendavid2020} which estimated prevalence less than 2\% in the county.  The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals. In these settings, SRS is likely to fail; however, there are potential solutions.  One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.

@@

\subsection*{Random sampling for doubly robust estimation}

@@

@@

>>>>

\section{Introduction}
The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of July 29th, 2021, over 196 million cases have been confirmed worldwide with 34.8 million cases and over 612 thousand confirmed deaths across the United States. This pandemic
has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.  Public databases~\citep{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.

@@

\subsection{Inverse-probability weighting approach}
\label{section:ipwapproach}

@@

\subsection{Application of SEIR model}
\label{section:modelbased}

Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates. Based on published age-specific IFRs~\cite{Levin2020}, IFR by age seems to closely follow a log-linear relationship. Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
\begin{table}[!th]
\begin{tabular}{c | c c c c c c c c}
Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
\end{tabular}
\caption{IFR by age-strata}
\label{tab:ifrperage}
\end{table}
The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.  See~\ref{app:modelbased} for additional details on parametrization and choice of priors. To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed. Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.  The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.

\begin{figure}[!th]
\centering
\includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
\caption{Posterior predictive check on aggregate daily death counts.}
\label{fig:deathsppc}
\end{figure}

@@

Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.  Figure~\ref{fig:air} presents the results. The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.  We conjecture this is due to the limited availability of symptom and other important covariates that may impact the likelihood an individual will get  tested. Our imputation strategies use available data to try and alleviate this issue; however, this suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.

While the model-based estimates appear more reasonable, note that the estimate on the last week of April is $0.1$\% which is a severe underestimate. Caution is therefore warranted when interpreting these estimates.  We know that the age-specific infection fatality rate is likely time-varying and may vary by other factors beyond age.  Therefore, improvements to these model-based predictions could be made by improving the IFR estimates; however, there is currently minimal publicly available data to do so.  Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.  The 95\% credible range for active infections on April 25th, for example, is $[XX,XX]$ with a posterior mean of $XX$.  Finally, the doubly robust estimates appear similar to the IPW estimate.

\begin{figure}[!th]
\centering
\includegraphics[width=.6\linewidth]{../figs/tv_air.png}
\caption{Active Infection Rate Estimates}
\label{fig:air}
\end{figure}

@@

There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty. This paper demonstrates how selection bias leads data analysts to feel certain about incorrect conclusions.  As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues. We end with a brief discussion of important related topics.

@@

Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We would argue that, after the current wave, understanding active infections is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

% \subsection*{Stratified sampling: improving precision in low-prevalence environments.}
% Random sampling is a powerful tool.  One difficulty with SRS is when prevalence is low.  Consider the Santa Clara study~\citep{Bendavid2020} which estimated prevalence less than 2\% in the county.  The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals. In these settings, SRS is likely to fail; however, there are potential solutions.  One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.

@@

\subsection*{Real-world implementation of the proposed method}

@@

Other biases exist and are incredible important when trying to estimate quantities such as the incubation period and mean serial interval.

These estimates could be severely biased due to selection bias and measurement error, but also suffer from

Nevertheless,
Length bias and measurement-error?

@@

\section{Model-based: Additional details}
\label{app:modelbased}
